Wormtable: A data structure for genome scale data

### 2013-07-02 docs/paper/paper.tex
Fixed author order and started reshuffle.

\title[Wormtable]{Wormtable: A data structure for genome scale tabular data}
\author[Kelleher \textit{et~al}]{Jerome Kelleher, Daniel L. Halligan 
    and Robert W. Ness }
\address{
University of Edinburgh,
King's Buildings,
West Mains Road,
EH9 3JT,
UK
}

@@

\begin{abstract}
\section{Summary:}
Biological research generates vast quantities of tabular data, which must 
be processed to answer biological questions. These tables are produced by 
programs and exported to plain text files, which are often tens of gigabytes
long. Researchers must process these files 
line-by-line, parsing values encoded as plain text into native types so that 
calculations may be done. This is an enormously inefficient process.
Furthermore, there is no simple means of indexing these files so that 
arbitrary data values within rows can be quickly found. We introduce a new 
data format and software library called Wormtable, which provides efficient 
access to tabular data using languages such as Python. Wormtable avoids the 
overhead of repeatedly parsing text, random access to rows, 
and also provides efficient indexing of arbitrary columns within these tables.
\section{Availability:}
Package available at 
\href{http://pypi.python.org/pypi/wormtable}{http://pypi.python.org/pypi/wormtable}

@@

\section{Introduction}
The large volumes of data that present day bioinformatics generates 
is well documented. The improvement in sequencing technologies has
lead to an exponential increase in the volume of data that is available 
to biologists. Perfectly standard studies require the storage and processing 
of hundreds of gigabytes of data, and this volume of data is only set to 
increase as sequencing technologies continue to improve. [\emph{This 
intro is intended to set the stage: there is shit loads of data out there,
and this is only set to increase}].

Despite the increasing importance of data processing, however, the methods 
used in Bioinformatics are crude and inefficient being almost entirely 
based around the use of text files. Plain text files have many advantages:
they are the most portable format available, and can be processed on 
any platform with the minimum of library dependencies. Text files also 
have many disadvantages however; the principle difficulties for our 
purposes here are, they are slow to parse and difficult to index.
% text files are also easy to read/comprehend and don't require specialist
% knowledge to process.
% I think learning a database syntax is a barrier to many researchers that are
% not trained bioinformaticians

One may, of course, load VCF data into a relational database, and 
use SQL to retrieve the records of interest. This approach has several 
disadvantages, however. Firstly, one must first examine the VCF 
header and construct an appropriate SQL schema to hold the columns 
of interest. It is feasible to do this for more recent versions of VCF, 
but is certainly not a trivial process, particularly if the schema 
must be portable across different relational database platforms.
Once the schema has been constructed, the data must be inserted into 
this database line-by-line by constructing SQL code, which is a 
time-consuming and error prone process. Even assuming that the data 
has been inserted into a table with an appropriate schema, there 
are still difficulties to be addressed. Maintaining a database 
server is a difficult task and when the volumes of data become
large, a specialised administrator is usually required to 
ensure the server works correctly.

Much of the time, a database server is far more complicated than
we require for large scale biological data. In the case of VCF,
the file is written once by an application such as [gatk? references?]
%Also samtools
and is not subsequently expected to change. Thus, storing this information 
in a relational database with its sophisticated concurrency control
and guarantees of consistency in the face of asynchronous updates
is entirely unnecessary. Even the client-server architecture of 
a typical relational database is wasteful in this case: protocol 
commands must be read and parsed, and responses must be encoded,
written to a socket and subsequently converted back to native types,
all of which constitutes a significant overhead when streaming large 
quantities of data. The limitations of relational database servers 
have been acknowledged for some time now, and has led to a 
variety of `NoSQL' database systems~\citep{m12}.

Databases do, however, have significant advantages over 
plain text files. Data is typically encoded in an efficient binary 
format, saving substantial space and giving better precision when 
dealing with floating point formats. Records are stored in a
B-tree type structure, leading to fast access times even 
over very large datasets. Indexing over columns within rows is 
also very powerful, allowing us to quickly retrieve 
rows with particular features. Ideally, we would like a data format
that combines these useful features of databases without the 
drawbacks of a relational database server outlined above.

Wormtable 
%... [Write Once Read Only Table] ...
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data with high-performance access to rows via indexes. Tables can
be either written directly via the software library, or converted
from existing formats via command line tools. Tables can be read 
concurrently by any number of processes, and are accessed by a 
simple API in Python.

% Also provided are scripts to import two common bioinformatic file standards - gff and vcf
% along with scripts to apply filters to these files?

\section{Wormtable}
To take advantage of the data processing capabilities of a database
without the overheads and disadvantages of a relational database 
server, Wormtable is built on top of Berkeley DB~\citep{obs99}.
Berkeley DB is an open source embedded database toolkit, that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}.

@@

\section{Examples}
Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples to 
illustrate the improvements over existing methods. To do this, 
we converted a VCF file with XXX rows from the YYY project 
into wormtable format. The (uncompressed) VCF file was XX
gigabytes, and the converted wormtable YY gigabytes, which required 
X hours.

>>>>

\title[Wormtable]{Wormtable: A data structure for genome scale data}
\author[Kelleher \textit{et~al}]{Jerome Kelleher, Robert W. Ness 
and Daniel L. Halligan}
\address{
University of Edinburgh,
King's Buildings,
West Mains Road,
EH9 3JT,
UK
}

@@

\begin{abstract}
\section{Summary:}
Biological research generates vast quantities of tabular data, which must 
be processed to test hypotheses. These tables are produced by 
programs and exported to plain text files, which are often very large. 
Researchers must process these files 
line-by-line, parsing values encoded as text into native types so that 
calculations may be done. This is an extremely inefficient process.
Furthermore, there is no simple means of indexing these files so that 
arbitrary data values within rows can be quickly found. We introduce a new 
data format and software library called wormtable, which provides efficient 
access to tabular data in Python. Wormtable stores data in a compact
binary format, provides random access to rows, and enables sophisticated
indexing on columns within these tables.
\section{Availability:}
Package available at 
\href{http://pypi.python.org/pypi/wormtable}{http://pypi.python.org/pypi/wormtable}

@@

\section{Introduction}
% Lots of data
The improvement in sequencing technologies has
lead to an exponential increase in the volume of data being managed 
by biologists. Standard studies require the storage and processing 
of hundreds of gigabytes of data, and this volume of data is only set to 
increase as sequencing technologies continue to improve.

% but we use text files
Despite the increasing importance of data processing, however, the methods 
used in Bioinformatics are almost entirely 
based around the use of text files. Plain text files have many advantages:
they are extremely portable and can be processed on 
any platform with the minimum of library dependencies. Text files also 
have many disadvantages however; the principle difficulties for our 
purposes here are, they are slow to parse and difficult to index.

% text files are also easy to read/comprehend and don't require specialist
% knowledge to process.
% I think learning a database syntax is a barrier to many researchers that are
% not trained bioinformaticians

% Databases solve these problems
The classical approach to working with large volumes of 
structured data is to use a database, which solve these 
problems very effectively.
Data is typically encoded in an efficient binary 
format, saving substantial space and giving better precision when 
dealing with floating point formats. Records are stored in a
B-tree type structure, leading to fast access times even 
over very large datasets. Indexing over columns within rows is 
also very powerful, allowing us to quickly retrieve 
rows with particular features.

% Databases but introduce several others. 
One may, of course, load textual data into a relational database, and 
use SQL to retrieve the records of interest. This approach has several 
disadvantages, however. Firstly, one must first examine the data 
construct an appropriate SQL schema to hold the columns 
of interest. It is sometimes feasible to do this automatically 
for data files, but is certainly not a trivial process, 
particularly if the schema 
must be portable across different relational database platforms.
Once the schema has been constructed, the data must be inserted into 
this database line-by-line by constructing SQL code, which is a 
time-consuming and error prone process. Even assuming that the data 
has been inserted into a table with an appropriate schema, there 
are still difficulties to be addressed. Maintaining a database 
server is a difficult task and when the volumes of data become
large, a specialised administrator is usually required to 
ensure the server works correctly.

% We don't need concurrency, etc, as it's always read-only.
Much of the time, a database server is far more complicated than
we require for large scale biological data. Data files 
are usually written once by an application 
and is not subsequently expected to change. Thus, storing this information 
in a relational database with its sophisticated concurrency control
and guarantees of consistency in the face of asynchronous updates
is entirely unnecessary. Even the client-server architecture of 
a typical relational database is wasteful in this case: protocol 
commands must be read and parsed, and responses must be encoded,
written to a socket and subsequently converted back to native types,
all of which constitutes a significant overhead when streaming large 
quantities of data. The limitations of relational database servers 
have been acknowledged for some time now, and has led to a 
variety of `NoSQL' database systems~\citep{m12}.

\section{Wormtable}

Wormtable (write-once read-many table) 
is a new data format and software library designed specifically
to deal with the challenges of data processing in bioinformatics.
It provides a portable, compact, read-only store for tabular
data with high-performance access to rows via indexes. Tables can
be either written directly via the software library, or converted
from existing formats via command line tools. Tables can be read 
concurrently by any number of processes, and are accessed by a 
simple API in Python.

To take advantage of the data processing capabilities of a database
without the overheads and disadvantages of a relational database 
server, Wormtable is built on top of Berkeley DB~\citep{obs99}.
Berkeley DB is an open source embedded database toolkit, that provides a
scalable key-value store. It is a mature and stable platform, 
and is currently the most widely deployed database toolkit in the 
world~\citep{sb12}.

@@

\section{Examples}

Wormtable is designed to make working with genome scale data more 
efficient, and in this section we describe some examples to 
illustrate the improvements over existing methods. To do this, 
we converted a VCF file with XXX rows from the YYY project 
into wormtable format. The (uncompressed) VCF file was XX
gigabytes, and the converted wormtable YY gigabytes, which required 
X hours.
